# Introduction

The following technical setup is needed to pursue a local simulation on your computer:  

-  A local installation of the R statistical programming language [@R-base] is 
   required.  
   The current version can be downloaded from [https://cran.r-studio.org](https://cran.r-project.org/). 
   
- In addition, we recommend the use of RStudio Desktop (an integrated development environment) 
  which provides an excellent user interface for working with R.  
  It is available from here: [https://www.rstudio.com/products/rstudio/download/](https://www.rstudio.com/products/rstudio/download/).   


Both R and RStudio (as well as further used R packages) are free software.  


To follow this tutorial, we assume basic familiarity with the use of the R language.  

## Required R packages

The following R packages are needed to run the simulation: 

  - {fidelitysim} [@R-fidelitysim] - holds packaged code for the simulation experiment    

It uses and builds upon: 

  - {samplingDataCRT} v1.0 [@R-samplingDataCRT]  - for sampling data matrices  
  
  - {lme4} [@R-lme4]  - for linear model estimation  
  
  - {ggplot2} [@R-ggplot2] - for visualizing the results  


You can install the necessary packages by:  

```{r, eval = F}
# to install {fidelitysim} we need the {remotes}-package
install.packages("remotes")

# then install {fidelitysim} from GitHub
remotes::install_github("INS-Basel/fidelitysim")
```

The other packages are installed from CRAN:  

```{r, eval = F}
# install packages from CRAN
install.packages(c("samplingDataCRT", "lme4", "ggplot2"))
```

Consequently, to start the calculation the package needs to be loaded by:  

```{r, eval = F, warning=F}
library(fidelitysim) 
```

## Other considerations

The local simulation can - depending on the number of clusters and the computational power - take up to 
10-12 hours when the commenly recommended number of iterations of 10,000 is used. We suggest the user to start own simulations with a small number and than increase, when it is clear the code is running.
